Towards an Ecological Trait-data Standard

### 2017-12-12 manuscript/manuscript.Rmd
add comments of Pete and Caterina (#26)

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. 
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97]. Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17]. 
--> 
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. 
To mobilize data for this enterprise, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. 
Nonetheless, individual researchers often neglect the benefit of centralised databases and, to cope with journals' and research funding agencies increasing demand for the publication of data without access restrictions [@swan12; @allison15], instead publish primary data on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, Dryad). Given this prospect, there will be no lack of data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold structure and vocabulary for describing trait-datasets compatible with existing ontologies. (5.) We provide a tool for an easy adoption of this structure and for the harmonisation of data from different sources.

With an easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold  trait data standardisation.

Despite the obvious benefit and grand prospects of functional approaches, an integration of several studies across taxa, trophic levels, or organisational scales remains difficult because of different organismal, methodological, and functional focus, which renders data on traits extremely heterogeneous.

Trait data have therefore been standardised into centralised databases for specific organism groups and research questions. These compiled data are either being published as static data tables or in the form of sophisticated web-services (see below for a review of current initiatives). Services attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, but more importantly, provide platforms for consensus building in methods. Furthermore, database services offer access control and data usage policies, to claim the rights of the original data providers, while simplifying data queries for synthesis researchers.

Nonetheless, open-access publications in low-threshold data repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation.

@@

After all, to mobilise the broad wealth of data into structured databases <!-- AO: Maybe better: formats-->, it requires a certain minimum of information from the original data providers. Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.  <!-- AO: The last paragraphs may need more sources/proofings?-->

@@

In this section, we review four kinds of initiatives of relevance for trait data standardisation: 
the first are projects that work with trait datasets that have been assembled out of a primary interest in the project context; 
the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and Thesauri; 
the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'semantic ontologies'.

@@

However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk. <!-- BKR: This might also result in data becoming no longer accessible (because the last computer on which the excel version required to read a file ran died -->
The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work. 
<!--To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or database management systems for scientific projects (e.g. BExIS, XX ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.--><!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information.

@@

Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories. 
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.

@@

@@

However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand in biodiversity data management for structured input from the research communities.

The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services. None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots --> 
The other reason is competence: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.

@@

Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one value that applies to the entity in question (i.e. to a species in the first case or to an occurrence in the latter case). The width of the table varies with the number of traits included. Merging data from different sources requires user-defined mapping and manual harmonization of these structures.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). 
<!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes. But not necessarily computational more effective at all. E.g. I start to mix up numeric values with descriptive values. I start to mix up different units. Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g. quality stuff. Ergo less effective from that perspective. But it has some advantages which may overrule dis. See below 
AO: For the whole proofing why to use such a long format I think that GBIF and DwC has already given some sentences. I strongly think their work should be considered and can be used as reference!
We can simply argue that a well established routine should now also executed on trait level!
Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.

--><!--
This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Long-table dataset purport multiple advantages for the data manipulation (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets. 
<!-- BKR: Is that really true? Of course, you have less problems on the schema or structure level. But, on the other hand, the differences (e.g. different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows). Thus, at least with respect to integration, you just move the problem from one level to the other.  To ease integration, one would need to agree on the entries in the features column
AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
Again: I think we find here papers regarding DWC and GBIF (and OBIS) as source of proofing the facts.
-->

The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID. Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig. 2*c*).

@@

@@

With this paper, we propose a versatile vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.

We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management. <!-- AO: Webservices!! ;  Really to be considered as standards? -->

@@

The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ... -->. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

@@

As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories. 
<!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
-->

To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. <!-- BKR: There is some grammatical error here; not sure what you try to say--> 
<!-- AG: This was a little bit confusing for me. In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon. On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name. In any case I would suggest to recommend to keep both the URI and the literal.-->
If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt') with a unified character encoding (e.g. UTF-8). <!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->

The metadata of any dataset that employs this data structure should refer to the online version as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v1.0/". In addition to the versioned online reference, the dataset should also cite this paper for an explanation of the rationale.

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09]. 
<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet. 
<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample. Here, location or measurement information are in the primary focus of the investigation (see vocabulary extensions below). 
A general definition, whether a column describes asset data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context. 
<!-- FDS: shorten this and refer to existing standards for database use. -->

@@

>>>>

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. 
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @jennings09; @simons16; @gossner16; @moretti13; @gamez-virues15; @kuhsel15;  @birkhofer14; @gravel16; @lajoie15], as well as macro-patterns of biodiversity at the global level [@diaz13; @diaz16; @reich99; @reich97]. Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].
Most importantly, the trait framework relates key ecosystem services for human well-being back to changes in biodiversity in consequence to global change or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @gossner16; @mangels17; @simons16; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15; @borschig13; @birkhofer15; @birkhofer17]. 
--> 
<!-- CP: I like very much this whole paragraph (both writing and content), but I feel that it’s a bit out of the topic. It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important). I give some suggestions on how to cut it and rather develop more the following one, which people are surely less aware of and which sets more the aim of the paper. I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.

That’s the structure I try to follow:
1. Traits are important in ecology, evolution and conservation (i.e. this paragraph)

2. To increase the scale of studies (spatial, temporal, taxonomical) we need to put together trait data from different sources. When we do this we can answer new questions and find generalities. So there is a growing interest in assembling trait data from different sources and taxa (e.g. give the number of papers using TRY).

3. There is a problem of standardization that makes it difficult to match trait datasets from different sources. Different reasons: no traits definitions, no common structure.

4. One solution are databases. But these are costly and depend on the will of few persons/funding. They also cover few taxonomic groups. Matching between different databases is still difficult because of different definitions and structures. Most data are still uploaded separately.
We need a solution to make their use easy.

But of course it’s just a suggestion, feel free to happily ignore it!

I would choose 2 or 3 (also for the previous), anyway it’s impossible to be exhaustive. Maybe put (e.g. ref1, ref2) so make clear that it’s just examples

-->
<!-- PM: Thuis is quite complex- e.g. ‘taxonomic impediment’etc. I actually think you can simplify this and built these references into the more general text above by adding a copule of more examples. I don’t think you need to say much about the merits of the trait based approach here- it’s covered elsewhere.

Huge number of references here- I think you can strip it down to a few key ones, and reviews. Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site. 
-->
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, and that continues to integrate more and more fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

However, it relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. <!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific? -->
To mobilize data for this enterprise <!-- PM: Please define what the exercise is- too vague at present
-->, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. <!-- PM: I think this is a key justification- perhaps say what happens if we don’t do this – non repeatable work, differing definitions of traits that may lead to scientific flaws, repetition of the painstaking task of processing trait data e.g. to compute species level means. 
-->
The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. <!-- CP: I felt that this needed a definition, not sure everybody know what an ontology is. Maybe we should guide more the reader that might not be familiar with the problems.
-->

Nonetheless, individual researchers often neglect the benefit of centralised databases <!-- PM: Explain what this benefit is. -->. Furthermore, many journals and research funding agencies require  data to be published without access restrictions [@swan12; @allison15]. However, this primary data is often posted on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, Dryad). The long-term outcome of this situation is an abundance of freely available data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from an pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice <!--PM: Current or your own new definition?--> for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold <!-- PM: What’s this? Please define
-->  structure and vocabulary for describing trait-datasets compatible with existing ontologies <!-- PM & CAT: define--> . (5.) Finally, we provide an R-package that assists in aply this structure to own data and for the harmonisation of data from distributed sources.

With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold <!-- PM:Again, I’m not sure what this is a low threshold of? --> trait data standardisation.

Despite the obvious benefit and great potential of functional trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used. All of these factors render trait data extremely heterogeneous.

As a response to these issues, trait data have been standardized in centralized databases for specific organism groups and research questions.
<!-- PM: (refs, also give an e.g. or two for some big ones like TRY- saw there was a new one for tropical mammals a couple of days ago too and a huge Chinese plants one- in )--> 
These compiled data are either being published as static data tables or in the form of sophisticated online portals like the plant trait database TRY (see below for a review of current initiatives). These portals attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, and importantly also, provide networks for consensus building in measurement methods. Furthermore, database services offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.

In contrast to standardized databases, open-access data publication in low-threshold repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org is becoming increasingly frequent. Such repositories foster decentralised data hosting with low thresholds regarding data standardisation and documentation.

@@

However, to mobilize the broad wealth of distributed data into structured databases <!-- AO: Maybe better: formats-->,  a certain minimum of information from original data providers is requrired, e.g. on the methodology and definitions of trait values or the covariates describing the ecological context of the trait observation. In order to get this accurate and well-defined input ecological researchers need to be aware of and understand basic principles of data standards and ontologies.  <!-- AO: The last paragraphs may need more sources/proofings?--><!-- PM: Literacy is quite a strong word- as illiterate is quite an insult. I think awareness is a problem too.  -->

 More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.

@@

In this section, we review four kinds of initiatives of relevance for trait data standardisation: 
the first are projects that work with trait datasets that have been assembled out of a primary interest in the project context; 
the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and Thesauri; 
the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.

@@

However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk. <!-- BKR: This might also result in data becoming no longer accessible (because the last computer on which the excel version required to read a file ran died -->
The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work.

As it stands, this lack of data standardization of low-threshold online repositories will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited.
<!--With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information. -->

@@

Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories. <!-- FDS: not an option for primary publication ...!-->
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains information on quantitative measurements or qualitative facts (i.e. trait values) describing the physical phenotypic characteristics relating to fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregating replicated measurements on multiple entities (Fig. 1*b*) or by estimation the means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in organisational scale -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or a higher-level taxon (e.g. a genus).

@@

For the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that intends to cover many potential use cases. This data structure is centered around a central data table that contains a single observation of a trait per row, linking it to a taxon and possibly also a single occurrence of the plant species (i.e. an individual specimen or 'entity'). 
<!-- FDS: enter comments of Jens Kattge here!-->

@@

However, the terms of the Darwin Core Standard and available extensions, as well as TraitBank, are responses to a demand from biodiversity data management for more structured input from the research community.

The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data to low-threshold  file-publishing services. <!-- PM: There is also little incentive for researchers to do this- why bother? Incentives are also needed. -->
None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. One reason for this is most certainly compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific --> 
The other reason is awareness and literaci: trait-data providers are not well trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off.

As it stands, this lack of data standardization will mean that that the compliation of data will be inefficient and that many published datasets will not be properly exploited. 
To cope with the trend towards distributed data sources, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g. BExIS, NEON ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into consistent and interlinked data infrastructures. One goal of such initiatives is to facilitate the publication and standardization of data for researchers, e.g. by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

@@

Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one value that applies to the entity in question (i.e. to a species in the first case or to an occurrence in the latter case). As the comlumn names and the width of the table varies with the number of traits included, merging data from different sources requires user-defined mapping and manual harmonization of these structures.

A computationally more effective method that allows for high flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16]. In these, each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). 
<!-- FDS: also refer to kattge 11 -->
<!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes. But not necessarily computational more effective at all. E.g. I start to mix up numeric values with descriptive values. I start to mix up different units. Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g. quality stuff. Ergo less effective from that perspective. But it has some advantages which may overrule dis. See below 
AO: For the whole proofing why to use such a long format I think that GBIF and DwC has already given some sentences. I strongly think their work should be considered and can be used as reference!
We can simply argue that a well established routine should now also executed on trait level!
Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.

--><!--
This allows repeated measurements on a single individual to be stored by linking the data from separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Long-table datasets purport multiple advantages for data manipulation (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. As long-table fromats draw from a defined set of columns,  merging of datasets is much easier. 
<!-- BKR: Is that really true? Of course, you have less problems on the schema or structure level. But, on the other hand, the differences (e.g. different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows). Thus, at least with respect to integration, you just move the problem from one level to the other.  To ease integration, one would need to agree on the entries in the features column
AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
Again: I think we find here papers regarding DWC and GBIF (and OBIS) as source of proofing the facts.
-->
<!--
The long-table format therefore is ideal for storing raw data of trait measurements in huge databases, and allows the entries to be linked via the respective unique identifiers to additional information. Each row of the dataset can therefore be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y' (Fig. 2*c*).

@@

<!-- PM: Can I suggest a glossary of terms – this would be really helpful for people who are not data management specialists

FDS: see Garnier et al. but may contain:

Term | Definition 
----- | -----
Terminology | 
Thesaurus | 
Ontology | 
semantic web |

-->

@@

With this paper, we propose a versatile vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based ecological research.
<!-- PM: It took a long time to get to this point- I suggest brininging in a section early on that describes what the paper will do as it’s not always clear where it is leading. --> 
<!-- FDS: point out focus on distributed low-threshold data, not full-fledged database initiatives. -->

In designing this structure, we drew on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We also paid particularl consideration the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of our proposed data structure with major trait databases and existing standards for biodiversity data management. <!-- AO: Webservices!! ;  Really to be considered as standards? -->

@@

We acknowledge that the vocabulary proposed will not capture the full complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ... -->. Therefore, we invite researchers to contribute to the next iterations of this standard terminology and develop own applications and ontologies.

<!-- PM: It seems that the object identifier for an individual is the key distinction here- this could be clearer.

Maybe you address this later but does your tool assign one of these- how to you make sure they are all unique if there are multiple users- e.g. if you pooled together two previously complied datasets which might contain some of the same individuals?
-->

@@

As described above, observation-level long-table formats are the most predictable and compatible format for the core content of trait datasets and should therefore also be used for standalone uploads on low-threshold repositories. 
<!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
-->
<!-- CP: Not sure this needs to be repeated -->

To enable a better standardisation of nomenclatures and values, the entries of the core dataset <!-- CP: Does this need to be defined --> should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. <!-- BKR: There is some grammatical error here; not sure what you try to say--> 
<!-- AG: This was a little bit confusing for me. In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon. On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name. In any case I would suggest to recommend to keep both the URI and the literal.-->
<!-- PM: '... failure to provide these URI for taxa will ...' In general some examples of benefits would be good -->
If data contain additional information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
<!-- PM: Not clear here -->
For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files  ('.csv' or '.txt') that are compatible with all computing platforms and internationalisation settings by applying a unified character encoding (e.g. UTF-8). <!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset, e.g."Schneider et al. 2017 Ecological Traitdata Standard v0.7, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/ETS/v0.6/". In addition to this versioned online reference, the dataset should also cite this paper for an explanation of the rationale.
<!-- cP: Maybe we should put in the appendix an example of dataset+metadata? This should probably go in the metadata section 
FDS: to consider!! -->

In very complex datasets, the repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09]. 
<!-- PM: Why do these avid this? -->
<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->

In some contexts, it may be safer to add such information directly as columns of the data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet. 
<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample. Here, location or measurement information are the primary focus of the investigation (see vocabulary extensions below). 
<!-- PM: I get lost here-->
A general recommendation, whether a column describes asset data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context. 
<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->

@@

<!-- PM: Can you say a little about what this vocab is and how it was chosen/defined? -->
